Migration Experience Trajectories (Working Title)

### 2022-05-14 preregistration/Preregistration3MPCA.Rmd
first draft

Two important developments, within the psychological sciences in general and in migration research in particular, have been (1) a shift towards more comprehensive assessment of human experiences and (2) a growing focus on longitudinal real-life data. Within the migration research field, recent reviews have been pushing for models that take into account affects, behaviors, cognitions, and desires when it comes to migrant adaptations [@Kreienkamp2022d; @Ward2019]. At the same time prominent figures in the field have called for studies that collect longitudinal [@Pettigrew1998; @Pettigrew2008; @Pettigrew2008b; @Pettigrew2011] and real-life experience-sampling data outside the lab [@MacInnis2015; @McKeown2017; @Dixon2005].

In the past, such data collections were often unfeasible because they were either physically impractical or too expensive. However, recent technological developments now allow us to easily collect large amounts of experience sampling data on mobile devices [e.g., @Keil2020] or using web-based applications [@Arslan2020]. And while generally speaking analytical methods for such, more complex, data have become more readily available [e.g., @ODonnell2021], it remains unclear how we should identify key developmental patterns --- especially across multiple variables at the same time (i.e., multiple aspects). In essence, the new extensive longitudinal data comes with new forms of heterogeneity when jointly considering differences between people, over time, and across variables. Yet, past analytical advances have almost exclusively pushed for stationary lagged regression models^[meaning models that assume stable means and variances over time; e.g., vector autoregression models, dynamic structural equation models, autoregressive integrated moving average models, and cross-lagged panel analyses] or basic trajectory models^[e.g., mixed effects models, spline regression models, and latent growth curve modeling.]. 
And while these two approaches are important for inferential model testing, both approaches ignore the more fundamental task of describing and unraveling the developmental data patterns.  
And while these two approaches are important for inferential model testing, we still miss methods for the more fundamental task of describing and unraveling the developmental data patterns.

Such descriptive methods are not merely important for hypothesis generation, but also for identifying and understanding adaptive and maladaptive patterns. We we need to identify which variables are most important in the adaptation of migrants, we need methods to identify clusters of similar individuals and developments, and we need to assess whether such clusters differ in key adaptation markers (including, well-being, intergroup anxiety, outgroup trust, or societal participation).

We recently conducted three experience sampling studies, following the migration experiences of recent migrants to the Netherlands. In this manuscript, we aim to assess the utility of an analytical methods that has recently received increasing attention for simultaneously decomposing/considering person-, variable-, and time heterogeneity. The proposed three mode principle component analysis is of particular interest because its approach mirrors the famous data box of participants, variables, and measurement occasions (Catell's data box) for understanding psychological data.

Problem: HOWEVER, ...

* unclear how do deal with this more complex data (many variables, persons, and time points).
  + many different variables could be important.
  + how to consider them jointly (e.g., VAR, DSEM, ARIMA, cross-lagged panel analysis for testing model predictions, limited to specified lag and mean stationarity; Trajectory focused: mixed effects model, spline regressions, latent growth curve modeling, limited: often univariate outcome). At the same time, analytical methods for such more complex data have become more readily available, making the analyses more approachable [@ODonnell2021]. 
* heterogeneity between people, over time, and across variables.
* **unclear how to identify core/important developments.** Especially across multiple variables at he same time (i.e., multiple aspects).
* unclear which variables, time scales, and methods are useful in practice.

Solution:

* 3MPCA uses the data cube

little data has thus far investigated the development of migration experiences and no research has assessed the co-development of multiple experience aspects. Yet, understanding how people differ in their migration trajectories, can be crucial in understanding adaptive and maladaptive patterns. Simultaneously clustering the person-, experience aspect-, and time level of migration experiences, using a three mode PCA, may allow to identify clusters of similar developments and whether these clusters differ in key adaptation markers (including, well-being, anxiety, trust, or societal inclusion).

@@

We do not have hypotheses in the traditional sense. Our analysis plan is based on the aim of describing and understanding a complex data set of extensive longitudinal data. We propose that the dimension reduction procedure we employ will identify meaningful patterns and developments within the data, which are useful to migration researchers and practitioners.

@@

For Study 1 (initial preliminary study) our target sample size is a sum of 1,000 daily diary measurements. As we expected a completion rate of around 80% we aimed to recruite 20 participants (20 participants X 50 measurements). We ultimately over-sampled slightly to compensate for potential drop outs given the length and intensity of the study. We then used several key relationships (relevant to the broader research team) to estimate sufficient power for a range of different analyses (i.e., using power simulations).

For Studies 2 and 3, our target sample size is a sum of 4,000 daily diary measurements. With 100% completion rate that would be archived with 67 participants (60 daily diary responses each). Given that we expect some incomplete daily diary measurements, we aimed to recruit 80 participants. For both studies we again over-sampled slightly to account for expected drop outs within each target population.

@@

Participants were recruited until the targeted participant number finish the pre-measurement. Invitations to complete additional daily diary measurements (in Study 3) were extended until participants chose to leave the study or at the most until two months (i.e., 64 days) after the initial entry survey (i.e., from the pre measurement survey).

@@

Given the circumstance that we utilize three independent experience sampling studies for our analyses, the variables are at times slightly different between studies. Additionally, given that the analysis we aim to undertake (i.e., 3MPCA) uses a large number of variables for the multiple imputations, dimension reduction, and correlational analyses, we will not list all items individually. Instead, we provide full variable information is available in the following data sheets: 'VariableSelectionDaily.xlsx' and 'VariableSelectionPrePost.xlsx'.

**Key variables**  
Most of the variables are identical across all studies, however some differ slightly in content or wording. However, for each study we aimed to collect the following types of variables. We collected substantially more motivational and affecive variables, as they have been understudied in the past.

@@

@@

**Auxiliary variables for multiple imputation**  
For the missing data imputation we follow the procedure reported by @Monden2015, and include any pre-, post-, or daily dairy variables that are significantly ($p$ < .01) and meaningfully ($r$ > .3) correlated with (1) the key variables or (2) missingness on the key variables. For an overview of the variables that qualify as auxiliary variables see the enclosed data sheets (see 'VariableSelectionDaily.xlsx' and 'VariableSelectionPrePost.xlsx'). Additional information about the multiple imputation procedure is also available at [Statistical models] 2.

@@

In our analysis, we follow the procedure outlined by @Monden2015

1. **Sample Selection.** Our sample selection consists of three main steps. We (1) aggregate the key variables over time to archive a reasonably interpretable number of time points and the remove a first proportion of missing data. We then address the selection of time point and participants. Given that multiple imputation procedures even work well with large proportions of missing data [assuming missingness at random; @Madley-Dowd2019], we decided on a general criterion of less than 45% missingness to balance sample size retention and bias in the multiple imputation model. Thus, we then (2) select the time points for which we have responses from 55% of the participants and (3) select participants who have responded to at least 55% of the surveys across the selected time range.
    1. Time point aggregation: Given that little data is available on the meaningful time scales of the selected psychological variables, we chose to choose the appropriate time scales using variance decomposition [e.g., see @Ram2014]. This is to say that we create multi-level unconditional means models (without predictors) that include possible nested time scales as levels. For example, assess the explained variances for measurements nested in days, nested in weeks, nested in months. 

@@

4. **Data preprocessing.** For the three mode PCA, we center (across participants but within time point) and to normalize (within variable but across time points) all key variables. The between-person centering, ensures that all variations are around the mean trend (which is removed by the centering). The normalization within variable are important for ensuring equal variances across variables, which ensures equal weighting of the variables in the 3MPCA. For methodological detail see [Transformations].

@@

9. **Component interpretation** To embed and interpret the resulting components (especially the person components), we calculate correlations of the component scores with key adaptation markers (including, well-being, anxiety, trust, or societal inclusion). If components remain unclear still we can additionally perform k-means clustering on the lower dimensional space to identify clearer groups which can be compared on the adaptation markers if necessary.

@@

For three mode PCA it is custom to center (across participants but within time point) and to normalize (within variable but across time points). This is done to create a meaningful zero value for each time point (to compare participants) and a variance of 1 across all time points of a variable (removing any overall trends).

We provide the main equations of the procedure below, where individual $i$ ($i = 1,\ ...,\ I$) reported variable $k$ ($k = 1,\ ...,\ K$) at time point $t$ ($t = 1,\ ...,\ T$), so that $I$=number of persons, $K$=number of items, $T$=number of time points are within the dataset. For further information as well as an example illustration see Supplemental Material 3 in @Monden2015.

@@

Many of the proposed procedures are more descriptive rather then inferential. Importantly, we will use amounts of variances explained as our main fit indices. 
If we perform inferential analyses, we will use the standard p<.05 criteria for determining whether the correlation and regression coefficients are statistically significant. We will place particular emphasis on effect sizes in our interpretations of the results.

>>>>

Recent reviews have called for more comprehensive assessment of human experiences and for more longitudinal real-life data, within the psychological sciences more broadly and in migration research in particular [e.g., @Kreienkamp2022d; @MacInnis2015; @McKeown2017; @Pettigrew2011; @Ward2019]. However, while generally speaking analytical methods for such, more complex, data have become more readily available [e.g., @ODonnell2021], it remains unclear how we should identify key developmental patterns --- especially across multiple variables at the same time.

In essence, the novel extensive longitudinal datasets come with new forms of heterogeneity, where we have to consider differences between people, over time, and across variables. Yet, past analytical advances have almost exclusively pushed for top-down inferential modeling procedures^[For example, stationary lagged regression models that assume stable means and variances over time (incl., vector autoregression models, dynamic structural equation models, autoregressive integrated moving average models, and cross-lagged panel analyses) or basic trajectory models (e.g., mixed effects models, spline regression models, and latent growth curve modeling).]. And while inferential model testing is certainly important, we still miss discussions of methods for the more fundamental task of describing, summarizing, and understanding the developmental data patterns.

As an example, little data has thus far investigated the development of migration experiences and no research has assessed the co-development of multiple experience aspects. Yet, understanding how people differ in their migration trajectories, can be crucial in understanding adaptive and maladaptive patterns. We we need to identify which variables are most important in the adaptation of migrants, we need methods to identify clusters of similar individuals and developments, and we need to assess whether such clusters differ in key adaptation markers (including, well-being, intergroup anxiety, outgroup trust, or societal participation). There is, thus, a clear need to assess the utility of analysis procedures focused on the description and understanding of complex dynamical data.

In this manuscript, we aim to assess the utility of one such promising analysis for the use of social psychological experience sampling data. Recent studies have laid out the potential effectiveness of dimension reduction procedures, which can address the new forms of person-, variable-, and time point heterogeneity jointly [i.e., three-mode principle component analyses, 3MPCA; e.g., @Monden2015]. We will, particularly, analyse three experience sampling studies, which followed the migration experiences of recent migrants to the Netherlands.

@@

We do not have hypotheses in the traditional sense. Our analysis plan is based on the aim of describing, summarizing, and understanding a complex data set of extensive longitudinal data. We propose that the dimension reduction procedure we employ will identify meaningful patterns and developments within the data, which are useful to migration researchers and practitioners.

@@

For Study 1 (initial preliminary study) our target sample size is a sum of 1,000 daily diary measurements. As we expected a completion rate of around 80% we aimed to recruite 20 participants (20 participants X 50 measurements). We further over-sampled slightly to compensate for potential drop outs given the length and intensity of the study.

We then used several key relationships (relevant to the broader research team) to estimate sufficient power for a range of different analyses (i.e., using power simulations). Based on these power simulations, for Studies 2 and 3, our target sample size was a sum of 4,000 daily diary measurements. With 100% completion rate that would be archived with 67 participants (60 daily diary responses each). Given that we expect some incomplete daily diary measurements, we aimed to recruit 80 participants. For both studies we again over-sampled slightly to account for expected drop outs within each target population.

@@

Participants were recruited until the targeted number of participant finished the pre-measurement. Invitations to complete additional daily diary measurements (in Study 3) were extended until participants chose to leave the study or at the most until two months (i.e., 64 days) after the initial entry survey (i.e., from the pre measurement survey).

@@

Given the circumstance that we utilize three independent experience sampling studies for our analyses, the variables are at times slightly different between studies. Additionally, given that the analysis we aim to undertake (i.e., 3MPCA) uses a large number of variables for the multiple imputations, dimension reduction, and correlational analyses, we will not list all items individually. Instead, we provide full variable information in the following data sheets: 'VariableSelectionDaily.xlsx' and 'VariableSelectionPrePost.xlsx'.

**Key variables**  
Most of the variables are identical across all studies, however some differ slightly in content or wording. However, for each study we aimed to collect the following types of variables.

@@

It should be noted, that we collected substantially more motivational and affecive variables, as they have been understudied in the past.

@@

**Auxiliary variables for multiple imputation**  
For the missing data imputation we follow the procedure reported by @Monden2015, and include any pre-, post-, or daily dairy variables that are significantly ($p$ < .01) and meaningfully ($r$ > .3) correlated with either the key variables or missingness on the key variables. For an overview of the variables that qualify as auxiliary variables see the enclosed data sheets (see 'VariableSelectionDaily.xlsx' and 'VariableSelectionPrePost.xlsx'). Additional information about the multiple imputation procedure is also available at [Statistical models] 2.

@@

For our analysis, we adapt the procedures outlined by @Monden2015:

1. **Sample Selection.** Our sample selection consists of three main steps. We (a) aggregate the key variables over time to archive a reasonably interpretable number of time points and the remove a first proportion of missing data. We then address the selection of time point and participants. Given that multiple imputation procedures even work well with large proportions of missing data [assuming missingness at random; @Madley-Dowd2019], we decided on a general criterion of less than 45% missingness to balance sample size retention and bias in the multiple imputation model. Thus, we then (b) select the time points for which we have responses from 55% of the participants and (c) select participants who have responded to at least 55% of the surveys across the selected time range.
    a. **Time point aggregation:** Given that little data is available on the meaningful time scales of the selected psychological variables, we chose to determine the appropriate time scales using variance decomposition [e.g., see @Ram2014]. This is to say that we create multi-level unconditional means models (without predictors) that include possible nested time scales as levels. We, for example, assess the explained variances for measurements nested in days, nested in weeks, nested in months. 

@@

4. **Data preprocessing.** For the three mode PCA, we center (across participants but within time point) and normalize (within variable but across time points) all key variables. The between-person centering, ensures that all variations are around the mean trend (which is removed by the centering). The normalization within variable are important for ensuring equal variances across variables, which ensures equal weighting of the variables in the 3MPCA. For methodological detail see [Transformations].

@@

9. **Component interpretation** To embed and interpret the resulting components (especially the person components), we calculate correlations of the component scores with key adaptation markers (including, well-being, anxiety, trust, or societal inclusion). If components remain unclear still we can additionally perform k-means clustering on the lower dimensional space to identify clearer groups, which can be compared on the adaptation markers if necessary.

@@

For three-mode PCA it is custom to center (across participants but within time point) and to normalize (within variable but across time points). This is done to create a meaningful zero value for each time point (to compare participants) and a variance of 1 across all time points of a variable (ensuring equal weightin of variables in the 3MPCA).

We provide the main equations of the procedure below, where individual $i$ ($i = 1,\ ...,\ I$) reported variable $k$ ($k = 1,\ ...,\ K$) at time point $t$ ($t = 1,\ ...,\ T$), so that $I$=number of persons, $K$=number of items, $T$=number of time points within the dataset. For further information as well as an example illustration see Supplemental Material 3 in @Monden2015.

@@

Many of the proposed procedures are descriptive rather then inferential. Importantly, we will use amounts of variances explained as our main fit indices. 
If we perform inferential analyses, we will use the standard p<.05 criteria for determining whether the correlation and regression coefficients are statistically significant. We will place particular emphasis on effect sizes in our interpretations of the results wherever necessary.
